---
title: "Introduction to programming in R"
author: "Julia Rühle & Leone Albinati"
date: "25th-27th of November 2024"
output:
  html_document:
    theme: cerulean
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
    highlight: tango
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_knit$set(root.dir = 'Y:/lalbinati/Rcourse_material_VHIO/')

library(ggplot2)
library(dplyr)
library(tidyverse)
library(patchwork)

```

# Course organization

> **Day 1 (27th of November):** 
>
* Introduction to R and Rstudio 
* R basics: objects, paths and functions
* The basis of reproducibility: the R script 
* Data types in R
* Data structures: vectors, matrices and dataframes
* Read and write files
* Libraries and packages

# Data types in R

Every object in R has a data type

- Numeric (integer or double): any number

- Character: text objects

- Logical: TRUE or FALSE

To check what is the data type of an object, we can use the function **mode()**

```{r}
# let's create different objects and check for their type
type1 <- 2
type2 <- "banana"
type3 <- "5"
type4 <- TRUE


# Can we multiply type1 and type3?
```

Checking for data types is important, especially when you will work with more complex data structures

# Data structures

The main data structures in R are:

- Vectors 

- Matrices

- Data Frames

There are also many other data structures (like factors - except if we need it for ggplot - and lists) that will not be discussed in this course.

&nbsp;
&nbsp;
&nbsp;
&nbsp;
&nbsp;
&nbsp;

> **Important note**
>
In this section, we will describe the main data structures used in R and we will show you how to handle and manipulate these structures using **basic R language**. As of today, an amazing set of packages to work and manipulate data structures has been developed, called **tidyverse**. Many codes are written using tidyverse and often ChatGPT and other language models suggest codes using tidyverse. Moreover, the well known package **ggplot** (the reason many of you are here?) is developed inside tidyverse. The way tidyverse handles data structures is different from how we do it with basic R. Still, **knowing basic R is important** (to properly know the language and to understand other people's scripts); therefore we decided to dedicate this session to learn how to handle data structures with basic R, and from tomorrow on we'll move to tidyverse.

&nbsp;
&nbsp;
&nbsp;
&nbsp;
&nbsp;
&nbsp;

## Vectors

Vectors are one-dimension sequences of data elements of the **same type**.

### Creating a vector

- To create a vector, we use the function **c()** - combine.
- The elements inside the c() have to be separated by commas.

```{r, include=TRUE, eval=FALSE}
# let's create a vector of numbers
my_first_vec <- c(1,3,5,7,9)
mode(my_first_vec)

# for numerical intervals, we can create a vector using the : (colon)
# the interval includes the extremes!
interval_vec <- 1:5

# let's create a character vector - each of the character elements has to be included in quotes!
# spaces between commas do not change the meaning
character_vec <- c("lion", "zebra", "elephant")
mode(character_vec)

# what happens if we don't put the quotes?
character_vec2 <- c(lion, zebra, elephant)

# what happens if we create a "mixed" vector?
mixed_vec <- c("lion", TRUE, 1)
mode(mixed_vec)

```


- tip1: to add the quotes you can select the elements and then press the quote key

- tip2: single quotes can also be used, but in R is preferable to use the double quotes

###  Combining vectors

Different vectors can be combined

```{r, include=TRUE, eval=FALSE}
# let's create the vector planets
gas_planets <- c("Jupiter", "Saturn", "Uranus", "Neptune")
terrestrial_planets <- c("Mercury", "Venus", "Earth", "Mars")
all_planets <- c(terrestrial_planets,gas_planets)

# we can also add single elements
really_all_planets <- c(all_planets,"Pluto")

# how many elements are there in the vector all_planets?
length(all_planets)
```


### Vector manipulation

Learning vector manipulation is essential in R because it forms the foundation for handling more complex data structures, such as data frames and matrices, which is crucial for data analysis and visualization.

- We can name vectors - might remove this one tbh

```{r, include=TRUE, eval=FALSE}
# let's create a vector of RNA counts
gene_counts <- c(125,36,42)

# let's name each element with the type of gene it belongs to
gene_types <- c("protein_coding","lncRNA","pseudogene")

# let's assign the gene type as names of the vector
names(gene_counts) <- gene_types
gene_counts

# it did not change mode - the names are not part of the vector
mode(gene_counts)
```

- We can **extract elements** from a vector using **their position (index)** inside the vector








